---
title: "Getting Started with keydrivers"
vignette: >
  %\VignetteIndexEntry{Getting Started with keydrivers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`kda()` runs up to eight variable importance methods in a single call and returns a combined importance matrix alongside per-method metadata. Variable types are detected automatically from R class, and each method selects the appropriate model for the outcome type.

This vignette uses the `mtcars` dataset (included in base R) to demonstrate three scenarios:

1. **Continuous outcome** — all eight methods on `mpg`
2. **Binary outcome** — key methods on a binarized version of `mpg`
3. **Ordinal outcome** — key methods on a discretized version of `mpg`

## Setup

```{r}
library(keydrivers)
data(mtcars)

x_vars <- c("disp", "hp", "wt", "drat")
```

`mtcars` contains performance and design data for 32 cars. We use fuel efficiency (`mpg`) as the outcome and four mechanical predictors as key drivers. Transmission type (`am`) will serve as a control variable in the continuous example.

---

## 1. Continuous Outcome

With a continuous outcome, all eight methods are available. Regression-based methods use linear regression and R².

```{r}
result <- kda(mpg ~ disp + hp + wt + drat | am,
              data       = mtcars,
              corr       = TRUE,
              beta       = TRUE,
              useful     = TRUE,
              jrw        = TRUE,
              shapley    = TRUE,
              shapex     = TRUE,
              randforest = TRUE,
              rf_params  = list(ntree = 1000),
              xgboost_   = TRUE,
              verbose    = FALSE)

print(result)
```

The importance matrix rows are key driver variables and columns are methods. `am` (transmission) is a control: it is held fixed in every regression but excluded from the bivariate correlations.

By default (`normalize = TRUE`) each method's scores sum to 100. A score of 32 for `wt` under Shapley means weight accounts for 32% of the full model's R². This common scale makes cross-method comparison meaningful and is what the plot uses on its x-axis. Pass `normalize = FALSE` to recover raw scores.

`wt` (weight) ranks first or second across nearly every method. `drat` (rear axle ratio) consistently ranks last. Cross-method agreement on rankings provides more confidence than any single method alone.

Each method accepts a corresponding `_params` list for tuning. `rf_params = list(ntree = 1000)` above increases the number of trees from the default 500 for more stable importance estimates. Other useful params: `xgb_params = list(nrounds = 200, eta = 0.1)`, `shapex_params = list(nsim = 100)`.

---

## 2. Binary Outcome

We binarize `mpg` at its median. Binary outcomes use logistic regression and McFadden pseudo-R² in place of linear regression and R².

```{r}
#| warning: false
mtcars$high_mpg <- as.integer(mtcars$mpg > median(mtcars$mpg))

result_bin <- kda(high_mpg ~ disp + hp + wt + drat,
                  data    = mtcars,
                  corr    = TRUE,
                  beta    = TRUE,
                  useful  = TRUE,
                  shapley = TRUE,
                  verbose = FALSE)

print(result_bin)
```

`kda()` detects `high_mpg` as binary (0/1 numeric) and routes each method to the correct model automatically — no arguments change from the continuous case. Note that `jrw` is unavailable for binary outcomes and would return NAs if requested. The Shapley pseudo-R² of 1 reflects near-perfect separation in this small dataset — the four predictors collectively predict `high_mpg` with very high accuracy, which is expected for `mtcars`.

---

## 3. Ordinal Outcome

We discretize `mpg` into three ordered tiers. Ordinal outcomes use proportional odds regression (`polr`) and McFadden pseudo-R².

```{r}
mtcars$mpg_tier <- cut(mtcars$mpg,
                       breaks        = quantile(mtcars$mpg, c(0, 1/3, 2/3, 1)),
                       labels        = c("low", "mid", "high"),
                       include.lowest = TRUE)
mtcars$mpg_tier <- factor(mtcars$mpg_tier, ordered = TRUE)

result_ord <- kda(mpg_tier ~ disp + hp + wt + drat,
                  data    = mtcars,
                  corr    = TRUE,
                  beta    = TRUE,
                  useful  = TRUE,
                  shapley = TRUE,
                  verbose = FALSE)

print(result_ord)
```

Ordinal x variables (if any) are automatically converted to integer ranks before being passed to regression models. Nominal (unordered factor) variables are blocked entirely — `kda()` stops with a clear message if one is supplied.

---

## 4. Bootstrap Confidence Intervals

`kda_boot()` wraps `kda()` and runs it on `B` bootstrap resamples to produce percentile confidence intervals for every importance score. All method flags and `_params` arguments are passed through unchanged.

```{r}
result_boot <- kda_boot(mpg ~ disp + hp + wt + drat,
                        data       = mtcars,
                        B          = 200,
                        conf_level = 0.95,
                        seed       = 42,
                        corr       = TRUE,
                        beta       = TRUE,
                        randforest = TRUE,
                        rf_params  = list(ntree = 500))

print(result_boot)
```

`print()` shows the full-data point estimates followed by the CIs. The `seed` argument makes results reproducible. For stochastic methods (RF, XGBoost, SHAP), reduce `ntree`, `nrounds`, or `nsim` when `B` is large to keep runtime manageable. The result object also exposes `result_boot$estimate` (the full-data `kda` object), `result_boot$ci_lower`, and `result_boot$ci_upper` for programmatic access.

---

## 5. Plotting

`plot()` dispatches on `kda` and `kda_boot` objects. Drivers are ordered from most to least important on the vertical axis; each method gets its own color; CI lines appear automatically for `kda_boot` results.

```{r}
# Point estimates from the continuous example
plot(result)
```

```{r}
# Bootstrap result with CIs — custom colors and background
plot(result_boot,
     colors   = c(correlation = "steelblue", beta = "tomato",
                  rf_importance = "forestgreen"),
     bg_color = "#f8f8f8")
```

The returned object is a `ggplot` and can be extended with any ggplot2 layer:

```{r}
#| eval: false
library(ggplot2)
plot(result) + ggtitle("Key Drivers of Fuel Efficiency")
```

---

## 6. Interpreting Results When Methods Disagree

Running eight methods and getting eight identical rankings would be suspicious. Some disagreement is normal and informative. The key is knowing what each method is measuring and what divergence between them is telling you.

### Why methods differ

Each method answers a subtly different question:

| Method | What it measures |
|--------|-----------------|
| Correlations | Raw bivariate association — each predictor vs. outcome in isolation |
| Betas | Partial association — each predictor controlling for all others simultaneously |
| Usefulness | Marginal R² contribution — fit loss when this variable is removed |
| JRW / Shapley | Fair partition of total R² across predictors, averaging over all orderings |
| RF / XGBoost | Non-linear and interaction-aware importance |

These are genuinely different questions. A variable can rank high on one and low on another without either answer being wrong.

### What divergence usually means

**Correlations high, Shapley/usefulness low**: the variable shares its predictive power with another predictor. Both are measuring the same underlying construct. JRW and Shapley will split the shared credit between them rather than double-counting.

**Betas in a different order than correlations**: a suppressor or multicollinearity pattern. A variable can have a modest bivariate correlation but become important once others are held constant (or vice versa). This is a real finding about the structure of the data, not a modeling artifact.

**Tree methods (RF, XGBoost) diverge from linear methods (Shapley, betas, usefulness)**: the relationship is probably non-linear or involves interactions between predictors. The linear methods fit a straight line and apportion its R²; the tree methods are picking up curvature or conditional structure that the line misses.

**Methods largely agree**: convergent evidence. Variables that rank consistently high across multiple methods are the most defensible findings.

### Investigating a linear/tree divergence

When Shapley and RF give meaningfully different rankings for a variable, work through these steps:

**1. Plot each predictor against the outcome with a smooth.**
A LOESS curve that departs from a straight line flags non-linearity directly.

```{r}
#| eval: false
library(ggplot2)

for (v in c("disp", "hp", "wt", "drat")) {
  p <- ggplot(mtcars, aes(x = .data[[v]], y = mpg)) +
    geom_point() +
    geom_smooth(method = "loess", se = TRUE) +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
    ggtitle(v)
  print(p)
}
```

**2. Partial dependence plots from the RF model.**
A PDP shows the RF's view of the functional form for each predictor, marginalizing over all others. A curved or step-shaped PDP confirms non-linearity. A flat PDP for a predictor the RF ranks highly suggests an interaction (the variable matters, but only conditionally).

```{r}
#| eval: false
library(randomForest)
library(pdp)

rf_model <- randomForest(mpg ~ disp + hp + wt + drat, data = mtcars, ntree = 1000)
partial(rf_model, pred.var = "wt", plot = TRUE, rug = TRUE)
```

**3. Compare linear model R² to a GAM.**
A GAM fits a smooth function to each predictor instead of a straight line. If the GAM's R² is meaningfully higher, the linear model is missing real structure and the Shapley values based on it are understating variables with curved relationships.

```{r}
#| eval: false
library(mgcv)

m_lm  <- lm(mpg ~ disp + hp + wt + drat, data = mtcars)
m_gam <- gam(mpg ~ s(disp) + s(hp) + s(wt) + s(drat), data = mtcars)

summary(m_lm)$r.squared
summary(m_gam)$r.sq   # materially higher = non-linearity matters
```

**4. Test for interactions.**
If the PDPs look roughly linear but RF still diverges, the likely culprit is an interaction — one variable's effect depends on the level of another. A 2D partial dependence plot reveals this: a non-flat surface indicates the two variables interact.

```{r}
#| eval: false
partial(rf_model, pred.var = c("wt", "hp"), plot = TRUE, chull = TRUE)

# Confirm with a formal test in the linear model
m_int <- lm(mpg ~ disp + hp + wt + drat + wt:hp, data = mtcars)
anova(m_lm, m_int)
```

---

## Formula Syntax

The pipe separator `|` partitions key drivers from control variables:

```r
kda(y ~ x1 + x2 + x3 | z1 + z2, data = df, ...)
```

Controls are included in every regression-based model but are not assigned importance scores. Their contribution to model fit is held constant before the key drivers are evaluated. Without controls, use the standard syntax:

```r
kda(y ~ x1 + x2 + x3, data = df, ...)
```
