---
title: "Getting Started with keydrivers"
vignette: >
  %\VignetteIndexEntry{Getting Started with keydrivers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

`kda()` runs up to eight variable importance methods in a single call and returns a combined importance matrix alongside per-method metadata. Variable types are detected automatically from R class, and each method selects the appropriate model for the outcome type.

This vignette uses the `mtcars` dataset (included in base R) to demonstrate three scenarios:

1. **Continuous outcome** — all eight methods on `mpg`
2. **Binary outcome** — key methods on a binarized version of `mpg`
3. **Ordinal outcome** — key methods on a discretized version of `mpg`

## Setup

```{r}
library(keydrivers)
data(mtcars)

x_vars <- c("disp", "hp", "wt", "drat")
```

`mtcars` contains performance and design data for 32 cars. We use fuel efficiency (`mpg`) as the outcome and four mechanical predictors as key drivers. Transmission type (`am`) will serve as a control variable in the continuous example.

---

## 1. Continuous Outcome

With a continuous outcome, all eight methods are available. Regression-based methods use linear regression and R².

```{r}
result <- kda(mpg ~ disp + hp + wt + drat | am,
              data       = mtcars,
              corr       = TRUE,
              beta       = TRUE,
              useful     = TRUE,
              jrw        = TRUE,
              shapley    = TRUE,
              shapex     = TRUE,
              randforest = TRUE,
              rf_params  = list(ntree = 1000),
              xgboost_   = TRUE,
              verbose    = FALSE)

print(result)
```

The importance matrix rows are key driver variables and columns are methods. `am` (transmission) is a control: it is held fixed in every regression but excluded from the bivariate correlations.

By default (`normalize = TRUE`) each method's scores sum to 100. A score of 32 for `wt` under Shapley means weight accounts for 32% of the full model's R². This common scale makes cross-method comparison meaningful and is what the plot uses on its x-axis. Pass `normalize = FALSE` to recover raw scores.

`wt` (weight) ranks first or second across nearly every method. `drat` (rear axle ratio) consistently ranks last. Cross-method agreement on rankings provides more confidence than any single method alone.

Each method accepts a corresponding `_params` list for tuning. `rf_params = list(ntree = 1000)` above increases the number of trees from the default 500 for more stable importance estimates. Other useful params: `xgb_params = list(nrounds = 200, eta = 0.1)`, `shapex_params = list(nsim = 100)`.

---

## 2. Binary Outcome

We binarize `mpg` at its median. Binary outcomes use logistic regression and McFadden pseudo-R² in place of linear regression and R².

```{r}
mtcars$high_mpg <- as.integer(mtcars$mpg > median(mtcars$mpg))

result_bin <- kda(high_mpg ~ disp + hp + wt + drat,
                  data    = mtcars,
                  corr    = TRUE,
                  beta    = TRUE,
                  useful  = TRUE,
                  shapley = TRUE,
                  verbose = FALSE)

print(result_bin)
```

`kda()` detects `high_mpg` as binary (0/1 numeric) and routes each method to the correct model automatically — no arguments change from the continuous case. Note that `jrw` is unavailable for binary outcomes and would return NAs if requested.

---

## 3. Ordinal Outcome

We discretize `mpg` into three ordered tiers. Ordinal outcomes use proportional odds regression (`polr`) and McFadden pseudo-R².

```{r}
mtcars$mpg_tier <- cut(mtcars$mpg,
                       breaks        = quantile(mtcars$mpg, c(0, 1/3, 2/3, 1)),
                       labels        = c("low", "mid", "high"),
                       include.lowest = TRUE)
mtcars$mpg_tier <- factor(mtcars$mpg_tier, ordered = TRUE)

result_ord <- kda(mpg_tier ~ disp + hp + wt + drat,
                  data    = mtcars,
                  corr    = TRUE,
                  beta    = TRUE,
                  useful  = TRUE,
                  shapley = TRUE,
                  verbose = FALSE)

print(result_ord)
```

Ordinal x variables (if any) are automatically converted to integer ranks before being passed to regression models. Nominal (unordered factor) variables are blocked entirely — `kda()` stops with a clear message if one is supplied.

---

## 4. Bootstrap Confidence Intervals

`kda_boot()` wraps `kda()` and runs it on `B` bootstrap resamples to produce percentile confidence intervals for every importance score. All method flags and `_params` arguments are passed through unchanged.

```{r}
result_boot <- kda_boot(mpg ~ disp + hp + wt + drat,
                        data       = mtcars,
                        B          = 200,
                        conf_level = 0.95,
                        seed       = 42,
                        corr       = TRUE,
                        beta       = TRUE,
                        randforest = TRUE,
                        rf_params  = list(ntree = 500))

print(result_boot)
```

`print()` shows the full-data point estimates followed by the CIs. The `seed` argument makes results reproducible. For stochastic methods (RF, XGBoost, SHAP), reduce `ntree`, `nrounds`, or `nsim` when `B` is large to keep runtime manageable. The result object also exposes `result_boot$estimate` (the full-data `kda` object), `result_boot$ci_lower`, and `result_boot$ci_upper` for programmatic access.

---

## 5. Plotting

`plot()` dispatches on `kda` and `kda_boot` objects. Drivers are ordered from most to least important on the vertical axis; each method gets its own color; CI lines appear automatically for `kda_boot` results.

```{r}
# Point estimates from the continuous example
plot(result)
```

```{r}
# Bootstrap result with CIs — custom colors and background
plot(result_boot,
     colors   = c(correlation = "steelblue", beta = "tomato",
                  rf_importance = "forestgreen"),
     bg_color = "#f8f8f8")
```

The returned object is a `ggplot` and can be extended with any ggplot2 layer:

```{r}
#| eval: false
library(ggplot2)
plot(result) + ggtitle("Key Drivers of Fuel Efficiency")
```

---

## Formula Syntax

The pipe separator `|` partitions key drivers from control variables:

```r
kda(y ~ x1 + x2 + x3 | z1 + z2, data = df, ...)
```

Controls are included in every regression-based model but are not assigned importance scores. Their contribution to model fit is held constant before the key drivers are evaluated. Without controls, use the standard syntax:

```r
kda(y ~ x1 + x2 + x3, data = df, ...)
```
